---
title: "Impute CMIPB 2020 and 2021 data using the MICE algorithm"
author: "Anna Konstorum (anna.konstorum@yale.edu)"
date: "3/8/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries
```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(jsonlite)
library(tibble)
library(UpSetR)
library(mice)
```


### Import Data (Visit 1)
Data source: [https://github.com/joreynajr/cmi-pb-multiomics/tree/main/results/main/cmi_pb_datasets/processed/harmonized]
Obtain files below from data source and save to local drive
```{r}
# data_dir is path to local drive where data is saved
setwd(data_dir)

abtiters.2020<-read.table('abtiters.mfi_normalized.2020.day0.pivoted.tsv',sep='\t',header=TRUE,stringsAsFactors=TRUE,row.names=1)
cytof.2020<-read.table('cytof.2020.day0.pivoted.tsv',sep='\t',header=TRUE,stringsAsFactors=TRUE,row.names=1)
olink.2020<-read.table('olink.2020.day0.pivoted.tsv',sep='\t',header=TRUE,stringsAsFactors=TRUE,row.names=1)
rnaseq.2020<-read.table('rnaseq.2020.day0.pivoted.tsv',sep='\t',header=TRUE,stringsAsFactors=TRUE,row.names=1)
meta.2020<-read.table('clinical_metadata.2020.tsv',sep='\t',header=TRUE,stringsAsFactors=TRUE,row.names=1)

abtiters.2021<-read.table('abtiters.mfi_normalized.2021.day0.pivoted.tsv',sep='\t',header=TRUE,stringsAsFactors=TRUE,row.names=1)
cytof.2021<-read.table('cytof.2021.day0.pivoted.tsv',sep='\t',header=TRUE,stringsAsFactors=TRUE,row.names=1)
olink.2021<-read.table('olink.2021.day0.pivoted.tsv',sep='\t',header=TRUE,stringsAsFactors=TRUE,row.names=1)
rnaseq.2021<-read.table('rnaseq.2021.day0.pivoted.tsv',sep='\t',header=TRUE,stringsAsFactors=TRUE,row.names=1)
meta.2021<-read.table('clinical_metadata.2021.tsv',sep='\t',header=TRUE,stringsAsFactors=TRUE,row.names=1)
```

#### Intersection of available 2020 data
```{r}
listInput = list(RNASeq = rownames(rnaseq.2020),Olink = rownames(olink.2020), 
                 CyTOF = rownames(cytof.2020), abTiter = rownames(abtiters.2020))

upset(fromList(listInput), order.by = "freq")
```

#### Heatmap of 2020 data availability (with RNASeq data subset for ease of visualization)
```{r}
rnaseq.2020_short<-rnaseq.2020[1:36,1:20]
X_joined_short <- dplyr::full_join(rownames_to_column(log2(as.data.frame(rnaseq.2020_short+1))),
                             rownames_to_column(cytof.2020), 
                             by = "rowname")
X_joined_short<-dplyr::full_join(X_joined_short,
                                 rownames_to_column(olink.2020),
                                 by = "rowname")
X_joined_short<-dplyr::full_join(X_joined_short,
                                 rownames_to_column(log2(abtiters.2020+1)),
                                 by = "rowname")

print(dim(X_joined_short))
X_mat <- t(as.matrix(dplyr::select(X_joined_short, -rowname)))

ggplot(reshape2::melt(X_mat)) +
  geom_tile(aes(x = Var1, y = Var2, fill = value, color = value), size = 1) +
  scale_fill_distiller(palette = "RdBu", direction = 1, na.value = "black") +
  scale_color_distiller(palette = "RdBu", direction = 1, na.value = "black", guide = FALSE) +
  ggtitle("RNA-Seq | CyTOF | OLINK") +
  xlab("Features") +
  ylab("Baseline Samples") +
  cowplot::theme_map() +
  theme(plot.title = element_text(hjust = .5, size = 14),
        axis.title = element_text(hjust = .5),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10, angle = 90))
```

#### Remove samples with no RNASeq data
```{r}
seq_u<-union(rownames(rnaseq.2020),rownames(olink.2020))
seq_u_c<-union(seq_u,rownames(cytof.2020))
samples_remove<-setdiff(rownames(abtiters.2020),seq_u_c)
abtiters.2020_filt<-subset(abtiters.2020,!(rownames(abtiters.2020) %in% samples_remove))

samples_remove<-setdiff(union(rownames(olink.2020),rownames(cytof.2020)),rownames(rnaseq.2020))
olink.2020_filt<-subset(olink.2020,!(rownames(olink.2020) %in% samples_remove))
cytof.2020_filt<-subset(cytof.2020,!(rownames(cytof.2020) %in% samples_remove))
abtiters.2020_filt<-subset(abtiters.2020_filt,!(rownames(abtiters.2020_filt) %in% samples_remove))

rnaseq.2020_short=rnaseq.2020[1:36,1:20]
X_joined <- dplyr::full_join(rownames_to_column(log2(rnaseq.2020_short+1)),
                             rownames_to_column(cytof.2020_filt),
                             by = "rowname")
print(dim(X_joined))
X_joined<-dplyr::full_join(X_joined,
                           rownames_to_column(olink.2020_filt),
                           by = "rowname")
X_joined<-dplyr::full_join(X_joined,
                           rownames_to_column(log2(abtiters.2020_filt+1)),
                           by = "rowname")
print(dim(X_joined))
            

X_mat <- t(as.matrix(dplyr::select(X_joined, -rowname)))

ggplot(reshape2::melt(X_mat)) +
  geom_tile(aes(x = Var1, y = Var2, fill = value, color = value), size = 1) +
  scale_fill_distiller(palette = "RdBu", direction = 1, na.value = "black") +
  scale_color_distiller(palette = "RdBu", direction = 1, na.value = "black", guide = FALSE) +
  ggtitle("RNASeq | CyTOF | OLINK | abtiters") +
  xlab("Features") +
  ylab("Baseline Samples") +
  cowplot::theme_map() +
  theme(plot.title = element_text(hjust = .5, size = 14),
        axis.title = element_text(hjust = .5),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10, angle = 90))
```

## Impute data using MICE algorithm
```{r}
seq_u<-union(rownames(rnaseq.2020),rownames(olink.2020))
seq_u_c<-union(seq_u,rownames(cytof.2020))
samples_remove<-setdiff(rownames(abtiters.2020),seq_u_c)
abtiters.2020_filt<-subset(abtiters.2020,!(rownames(abtiters.2020) %in% samples_remove))

samples_remove<-setdiff(union(rownames(olink.2020),rownames(cytof.2020)),rownames(rnaseq.2020))
olink.2020_filt<-subset(olink.2020,!(rownames(olink.2020) %in% samples_remove))
cytof.2020_filt<-subset(cytof.2020,!(rownames(cytof.2020) %in% samples_remove))
abtiters.2020_filt<-subset(abtiters.2020_filt,!(rownames(abtiters.2020_filt) %in% samples_remove))

X_joined <- dplyr::full_join(rownames_to_column(log2(rnaseq.2020+1)),
                             rownames_to_column(cytof.2020_filt),
                             by = "rowname")
print(dim(X_joined))
X_joined<-dplyr::full_join(X_joined,
                           rownames_to_column(olink.2020_filt),
                           by = "rowname")
X_joined<-dplyr::full_join(X_joined,
                           rownames_to_column(log2(abtiters.2020_filt+1)),
                           by = "rowname")
print(dim(X_joined))
            

X_mat <- t(as.matrix(dplyr::select(X_joined, -rowname)))

imputed_Data <- mice(X_mat, m=1, maxit = 25, method = 'pmm', seed = 500)
```

#### Save 2020 imputed data into dataframes
```{r}
X_mat_complete <- complete(imputed_Data,1)
colnames(X_mat_complete) <- X_joined$rowname

X_mat_complete_df <- as.data.frame(t(X_mat_complete))
X_mat_complete_df_2020<-X_mat_complete_df

# Split up into datasets:
rnaseq_baseline_mat_imputed_20 <- X_mat_complete_df[colnames(rnaseq.2020)]
cytof_baseline_mat_imputed_20 <- X_mat_complete_df[colnames(cytof.2020_filt)]
olink_baseline_mat_imputed_20 <- X_mat_complete_df[colnames(olink.2020_filt)]
abtiters_baseline_mat_imputed_20 <- X_mat_complete_df[colnames(abtiters.2020_filt)]
```


#### Visualize imputed data
```{r}

X_joined <- dplyr::full_join(rownames_to_column(rnaseq_baseline_mat_imputed_20[,1:20]),
                             rownames_to_column(cytof_baseline_mat_imputed_20),
                             by = "rowname")
X_joined<-dplyr::full_join(X_joined,
                           rownames_to_column(olink_baseline_mat_imputed_20),
                           by = "rowname")
X_joined<-dplyr::full_join(X_joined,
                           rownames_to_column(abtiters_baseline_mat_imputed_20),
                           by = "rowname")
            

X_mat <- t(as.matrix(dplyr::select(X_joined, -rowname)))

ggplot(reshape2::melt(X_mat)) +
  geom_tile(aes(x = Var1, y = Var2, fill = value, color = value), size = 1) +
  scale_fill_distiller(palette = "RdBu", direction = 1, na.value = "black") +
  scale_color_distiller(palette = "RdBu", direction = 1, na.value = "black", guide = FALSE) +
  ggtitle("RNASeq | CyTOF | OLINK | abtiters") +
  xlab("Features") +
  ylab("Baseline Samples") +
  cowplot::theme_map() +
  theme(plot.title = element_text(hjust = .5, size = 14),
        axis.title = element_text(hjust = .5),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10, angle = 90))
```

#### Save imputed data
```{r}
# imputed_dir: Local directory to save imputed data to
setwd(imputed_dir)
write.csv(rnaseq_baseline_mat_imputed_20,'rnaseq_baseline_mat_imputed_20_051022.csv')
write.csv(cytof_baseline_mat_imputed_20,'cytof_baseline_mat_imputed_20_051022.csv')
write.csv(olink_baseline_mat_imputed_20,'olink_baseline_mat_imputed_20_051022.csv')
write.csv(abtiters_baseline_mat_imputed_20,'abtiters_baseline_mat_imputed_20_051022.csv')

```

#### Repeat with 2021 data
```{r}
listInput = list(RNASeq = rownames(rnaseq.2021),Olink = rownames(olink.2021), 
                 CyTOF = rownames(cytof.2021), abtiters=  rownames(abtiters.2021))

upset(fromList(listInput), order.by = "freq")
```

#### Visualize un-imputed data
```{r}
rnaseq.2021_short<-rnaseq.2021[1:36,1:20]
X_joined <- dplyr::full_join(rownames_to_column(log2(as.data.frame(rnaseq.2021_short+1))),
                             rownames_to_column(cytof.2021), 
                             by = "rowname")
X_joined<-dplyr::full_join(X_joined,
                                 rownames_to_column(olink.2021),
                                 by = "rowname")
X_joined<-dplyr::full_join(X_joined,
                                 rownames_to_column(log2(abtiters.2021+1)),
                                 by = "rowname")
X_mat <- t(as.matrix(dplyr::select(X_joined, -rowname)))

ggplot(reshape2::melt(X_mat)) +
  geom_tile(aes(x = Var1, y = Var2, fill = value, color = value), size = 1) +
  scale_fill_distiller(palette = "RdBu", direction = 1, na.value = "black") +
  scale_color_distiller(palette = "RdBu", direction = 1, na.value = "black", guide = FALSE) +
  ggtitle("RNA-Seq | CyTOF | OLINK | abtiter") +
  xlab("Features") +
  ylab("Baseline Samples") +
  cowplot::theme_map() +
  theme(plot.title = element_text(hjust = .5, size = 14),
        axis.title = element_text(hjust = .5),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10, angle = 90))
```

#### Impute datasets
```{r}
X_joined <- dplyr::full_join(rownames_to_column(log2(as.data.frame(rnaseq.2021+1))),
                             rownames_to_column(cytof.2021), 
                             by = "rowname")
X_joined<-dplyr::full_join(X_joined,
                                 rownames_to_column(olink.2021),
                                 by = "rowname")
X_joined<-dplyr::full_join(X_joined,
                                 rownames_to_column(log2(abtiters.2021+1)),
                                 by = "rowname")
X_mat <- t(as.matrix(dplyr::select(X_joined, -rowname)))
imputed_Data_2021 <- mice(X_mat, m=1, maxit = 25, method = 'pmm', seed = 500)
```

#### Save 2021 imputed data into dataframes
```{r}
X_mat_complete <- complete(imputed_Data_2021,1)
colnames(X_mat_complete) <- X_joined$rowname

## visualize here

X_mat_complete_df <- as.data.frame(t(X_mat_complete))
#rownames(X_mat_complete_df) <- X_joined$rowname
X_mat_complete_df_2021<-X_mat_complete_df

# Split up into datasets:
rnaseq_baseline_mat_imputed_21 <- X_mat_complete_df_2021[colnames(rnaseq.2021)]
cytof_baseline_mat_imputed_21 <- X_mat_complete_df_2021[colnames(cytof.2021)]
olink_baseline_mat_imputed_21 <- X_mat_complete_df_2021[colnames(olink.2021)]
abtiters_baseline_mat_imputed_21 <- X_mat_complete_df[colnames(abtiters.2021)]
```

#### Save imputed data
```{r}
setwd(imputed_dir)
write.csv(rnaseq_baseline_mat_imputed_21,'rnaseq_baseline_mat_imputed_21_051022.csv')
write.csv(cytof_baseline_mat_imputed_21,'cytof_baseline_mat_imputed_21_051022.csv')
write.csv(olink_baseline_mat_imputed_21,'olink_baseline_mat_imputed_21_051022.csv')
write.csv(abtiters_baseline_mat_imputed_21,'abtiters_baseline_mat_imputed_21_051022.csv')
```

#### Visualize imputed data
```{r}
X_joined <- dplyr::full_join(rownames_to_column(rnaseq_baseline_mat_imputed_21[,1:20]),
                             rownames_to_column(cytof_baseline_mat_imputed_21),
                             by = "rowname")
X_joined<-dplyr::full_join(X_joined,
                           rownames_to_column(olink_baseline_mat_imputed_21),
                           by = "rowname")
X_joined<-dplyr::full_join(X_joined,
                           rownames_to_column(abtiters_baseline_mat_imputed_21),
                           by = "rowname")
            

X_mat <- t(as.matrix(dplyr::select(X_joined, -rowname)))

ggplot(reshape2::melt(X_mat)) +
  geom_tile(aes(x = Var1, y = Var2, fill = value, color = value), size = 1) +
  scale_fill_distiller(palette = "RdBu", direction = 1, na.value = "black") +
  scale_color_distiller(palette = "RdBu", direction = 1, na.value = "black", guide = FALSE) +
  ggtitle("RNASeq | CyTOF | OLINK | abtiters") +
  xlab("Features") +
  ylab("Baseline Samples") +
  cowplot::theme_map() +
  theme(plot.title = element_text(hjust = .5, size = 14),
        axis.title = element_text(hjust = .5),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10, angle = 90))
```